---
title: A Short Story of COVID-19 in Singapore
author: Brendi Ang
date: '2020-09-01'
slug: covid-19-in-singapore
categories:
  - R
tags:
  - R Markdown
  - Academic
subtitle: ''
summary: ''
authors: []
lastmod: '2020-09-01T22:09:55+10:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
bibliography: [references.bib]
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE)

# Load library
library(tidyverse)
library(coronavirus) # devtools::install_github("RamiKrispin/coronavirus")
library(DT)
library(janitor)
library(lubridate)
library(plotly)
library(readxl)
library(stringr)
library(timevis)
library(widgetframe)
library(zoo)
library(scales)
```

Undoubtedly, The COVID-19 pandemic will have altered personal lives, be it enormous or minor. There will be many stories told of how they 'survived' this pandemic down the road, having to fight for toilet papers and socially distancing from people outside. As a student, I'd like to add that group assignments studying online is a chore when you can't have face-to-face conversations. 

When I think about the detrimental implications of COVID-19, jobs and economy come to my mind. In Singapore, this pandemic has caused significant downturns for major industries, resulting to an the unemployment rate of 2.9% (@sgjobs), an all-time high since the global financial crisis.

My first blog post will feature a short story of the COVID-19 situation in my home country, Singapore. I will first study the trends in COVID-19 cases in Singapore, where I will break down the observations into 4 distinct components relating to the lockdown restrictions. Next, I will investigate the broad economical impact of COVID-19 on the different sectors of Singapore's economy. Notably, I included an interactive table for both of these sections as I thought it can be used in conjunction with figures to articulate the findings better.

## Getting the Data

To analyse the COVID-19 cases, I extracted the daily summary of Singapore's COVID-19 cases in the `coronavirus` package. This package contains the raw data in a tidy format, summarising the daily cases by state/province and is readily available on [GitHub](https://github.com/RamiKrispin/coronavirus). Each observation represents the total number of cases recorded in the state/province, where each cases are identified by their case type, spanning from 2020-01-23. It was really convenient to use this data set as minimal data wrangling had to be done.

The main variables that were utilised from the data set were:  
- `date`: Date in YYYY-MM-DD format
- `type`: Type of cases (confirmed, death or recovered)
- `cases`: Number of cases on a given date  

Next, to study the economical impact of COVID-19, I download the quarterly Gross Domestic Product(GDP) at Current Prices" from the [Department of Statistics Singapore website](https://www.tablebuilder.singstat.gov.sg/publicfacing/createDataTable.action?refId=16059) (@sggdp). In terms of the data cleaning process, the data was transposed before I removed unnecessary variables and renamed variables names such that they are consistent and comprising of only characters, numbers, lower case letters, with spaces denoted with underscores.

```{r read-in-data}
# From coronavirus package: long format of covid cases
sg_long <- coronavirus %>% 
  filter(country == "Singapore",
         date != "2020-01-22") %>% # Show first case recorded onwards
  select(-c(province, lat, long)) # No province in Singapore
```

```{r wrangle-gdp, include = FALSE}
# read-in GDP data
sg_gdp <- read_excel("OutputFile.xlsx")

sg_gdp <- sg_gdp %>%
  filter(!is.na(Variables)) %>%
  slice(1:(n()- 3))  # Remove last 3 rows relating to contact info

# Transpose data
sg_gdp <- as.data.frame(t(as.matrix(sg_gdp))) 

colnames(sg_gdp) <- sg_gdp[1,] 

sg_gdp <- sg_gdp %>% #
  slice(-1)

# Add Year variable& tidy data set 
sg_gdp <- sg_gdp %>%
  rownames_to_column('Year') %>%
  janitor::clean_names() # Remove spacings in variable names
```

### COVID-19 Cases in Singapore

```{r sg-phases}
# Duration of each government intervention 
sg_long <- sg_long %>%
  mutate(phase =
           case_when(
          between(date, as.Date("2020-01-22"), as.Date("2020-04-06")) ~ "No lockdown",
          between(date, as.Date("2020-04-07"), as.Date("2020-06-01")) ~ "First Lockdown",
          between(date, as.Date("2020-06-02"), as.Date("2020-06-17")) ~ "Phase 1 (Safe Reopening)",
          between(date, as.Date("2020-06-18"), as.Date("2020-09-01")) ~ "Phase 2 (Safe Transition)")
         )

# Compute 3-day Rolling average 
sg_long <- sg_long %>%
  group_by(type) %>%
  # 3-day rolling average
  mutate(rolling_avg_cases = round(zoo::rollmean(cases, k = 3, fill = 0), digits = 2), 
         cum_cases = cumsum(cases)) %>% # cumulative cases for each case type
  relocate(phase, .after = cum_cases)

# daily changes in cases
sg_long <- sg_long %>%
  mutate(lag_cases = lag(cum_cases, k = 1)) %>%
  mutate(daily_prop_change = (cum_cases - lag_cases)/lag_cases)

# Start and end date for each government intervention
lockdown_dur <- sg_long %>% 
  group_by(phase) %>%
  summarise(start = min(date),
            end = max(date))
```

```{r static-cases, fig.height = 8, fig.align = 'left'}
# Static plot: 3-day rolling averages
ggplot(sg_long) +
  # plot covid cases
  geom_line(aes(x = date,
                y = rolling_avg_cases,
                colour = type),
            alpha = 0.8) +
  theme_bw() +
  scale_colour_manual(values = c("#bc3a3a", "#00468b" ,"#468b00")) +
  theme(legend.position = "bottom") +
  scale_x_date(date_breaks = "1 week", 
               date_labels = "%b %d") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid.minor.x = element_blank()) +
  # Colour each phase of government intervention 
  geom_rect(data = lockdown_dur,
            aes(xmin = start, 
                xmax = end,
                ymin = 0, 
                ymax = 1500,
                fill = phase),
            alpha = 0.15) +
  ggtitle("Number of confirmed, death and recovered COVID-19 Cases") +
  scale_fill_manual(values = c('#c74848', '#228b57', '#72bcd4', '#a284e0')) +
  theme(legend.position = "bottom",
        legend.direction = "vertical")
```

The figure displays the number of new cases in Singapore daily by each case type. The the x-axis represents each date while the y-axis shows the number of new cases on a given day. To investigate the trends further, I have broken down the dates into four components, each representing the lockdown phase.

A marked escalation of new COVID-19 cases can be observed in early April, where the first lockdown was implemented, 7 April (*reference*). At a glance, it seems that Singapore was rather too late in tightening its restrictions and this could be avoided. Based on the *table reference*, Singapore's cases increased by a staggering `r ((1136.61-94.41)/94.41)*100`% from the first week to the last week of April although lockdown and social distancing protocols were in effect. 

Upon further investigation, this increase can be attributed to the surge in infections among migrant worker population staying in dormitories (@sgdorm). The impact of COVID-19 revealed that thousands of low-skilled migrant workers in Singapore lived in poor housing conditions. For instance, many migrant workers had "20 people sharing a single room" (@sgdorm2) where social distancing was too difficult to practice. In a pandemic, these living environment could serve as flash points for a further spread, as seen in the figure (@Koh634). As of 6 May, there were 17,758 confirmed COVID-19 cases among dormitory workers (approximately 90% of 20,000 nationally confirmed cases) (@Koh634). 

Although the confirmed cases were higher in Singapore, the fatality rate of COVID-19 cases were insignificant as compared to the new cases, staying relatively constant at 0. Is it a matter of chance where the people who were infected were healthy? At first, studies indicated that this coincidence may be due to the prevalence of testing as the Singapore's screening program comprises of young and even asymptotic people (@sgdeaths). Of course, the efforts of first line workers should never downplayed; And these assertions were quickly invalidated as research proved that salient factors other other than patient characteristics that determine Covid-19 mortality, such as government response are more prominent (liang2020covid). 

At the heart of Singapore’s operational capacity is its healthcare system, ranked among the top in the world, the healthcare system has been described as ‘high quality, low cost’. In 2019 itself,The National Centre for Infectious diseases (NCID) installed bed purpose built facility that were designed to manage an outbreak on the scale of SARS, allowing for excess capacity of patients (liang2020covid). 

Not to mention, the government was largely successful in contact tracing, and instilled a hefty penalty for the defiance of lockdown restrictions - fined up to $9,500 and/or up to six months of imprisonment. Therefore, the gradual decrease in the number of cases as seen on the figure may be a results of these stringent lockdown restrictions.

## Interactive table: Covid-19 Cases in Singapore

```{r}
# Replace inf, NaN and NA values with zero
sg_long$daily_prop_change[!is.finite(sg_long$daily_prop_change)] <- 0

# sg_long$daily_prop_change <- scales::percent(round(sg_long$daily_prop_change), digits = 4)

sg_long <- sg_long %>%
  mutate(daily_prop_change = round(daily_prop_change, digits = 2)) %>%
         # daily_prop_change = scales::percent(daily_prop_change)) %>%
  mutate(type = str_to_title(type)) # Capitalise first letter

# Rearrange table
tb1 <- sg_long %>%
  select(-c(country, lag_cases)) %>%
  relocate(phase, .after = daily_prop_change) %>%
  relocate(rolling_avg_cases, .before = daily_prop_change)
```

```{r}
tb1_interactive <- DT::datatable(tb1,
              colnames = c('Date Recorded', 
                           'Case Type', 
                           'Number of Cases',
                           'Cumulative Cases',
                           '3-day Rolling Average Cases',
                           'Daily Change in Cases (%)',
                           'Phase'
                           ),
              filter = 'top',
              rownames = FALSE,
              extensions = "Scroller",
              options = list(pageLength = 10,
                             autoWidth = TRUE,
                             scrollY = "250px")) %>%
  formatPercentage('daily_prop_change', digits = 2) %>% # Format proportion to percentage
  formatStyle(c('daily_prop_change'), `text-align` = 'right') %>% # Align values to the right 
  # Colour background of each lockdown stage
  formatStyle('phase',
              backgroundColor = styleEqual(
                unique(tb1$phase), c('#df9797', '#bfd9bf', '#d4ebf2', '#869cda'))
              ) 

# Frame table
widgetframe::frameWidget(tb1_interactive)
```


## COVID-19 impact on Singapore's economy

```{r wrangle-sgp-gdp}
# Convert character variables to numeric
sg_gdp <- sg_gdp %>%
  mutate(across(gdp_at_current_market_prices:add_taxes_on_products, as.numeric))

# Change quarter to date
sg_gdp <- sg_gdp %>%
  separate(year, into = c("year", "qtr"), sep = " ") %>% # separate year and quarter
  mutate(qtr = parse_number(qtr)) %>% # parse number for quarter 
  mutate(qtr = case_when(
    qtr == 1 ~ "03-31",
    qtr == 2 ~ "06-30",
    qtr == 3 ~ "09-30",
    qtr == 4 ~ "12-31"
  ))

sg_gdp <- sg_gdp %>%
  rowwise() %>%
  mutate(date = paste0(year, "-", qtr)) %>%
  mutate(date = as.Date(date)) %>%
  relocate(date, .before = gdp_at_current_market_prices) %>%
  select(-c(year, qtr))
```

```{r gdp-long}
# Convert GDP dataset to long format for plotting

# Create subgroups for the sectors
sg_gdp_long <- sg_gdp %>%
  # Keep aggregated scores together so it doesn't get 'pivoted'
  relocate(services_producing_industries, .before = manufacturing) %>% 
  pivot_longer(cols = manufacturing:ownership_of_dwellings,
               names_to = "Sector",
               values_to = "Sector_GDP") %>%
  # Categorise each sector according to their respective industry
  mutate(industry = case_when(
    Sector %in% c("manufacturing", "construction", "utilities", "other_goods_industries") ~ "goods",
    Sector %in% "ownership_of_dwellings" ~ "ownership",
    TRUE ~ "services")
    ) %>%
  # Capitalise first words of character string
  mutate(industry = str_to_title(industry),
         Sector = str_to_title(Sector)) 

# Compute GDP before taxes, to be consistent with sectors           
sg_gdp_long <- sg_gdp_long %>%
  mutate(gdp_at_current_market_prices = gdp_at_current_market_prices - add_taxes_on_products) %>%
  select(-c(add_taxes_on_products, gross_value_added_at_basic_prices)) %>%
  mutate(across(where(is.character), as.factor)) # Convert character strings to factor
```

```{r}
# Compute summary statistics at industry level
ind_summary <- sg_gdp_long %>%
  group_by(date, industry) %>%
  summarise(total = sum(Sector_GDP)) %>%
  mutate(month = month(date)) %>%
  # Quarter of each year
  mutate(qtr = month/3) %>%
  select(-month)
```

```{r}
# Find relative change of GDP for each industry per year
ind_change <- ind_summary %>%
  group_by(industry) %>%
  mutate(lag_total = lag(total, k = 1)) %>%
  mutate(pct_change = round((total-lag_total)/lag_total*100, digits = 4)
  ) 

p3 <- ggplot(ind_change) +
  geom_line(aes(x = date,
                y = pct_change,
                colour = industry),
            alpha = 0.8) +
  geom_hline(aes(yintercept = 0),
             alpha = 0.2) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~ industry,
             ncol = 1,
             strip.position = "top") +
  labs(y = "Quarterly change in GDP (%) for each Industry",
       x = "Date")

p3
```

As we understand the COVID-19 situation in Singapore better, we delve into investigating which sector are most adversely affected by the pandemic.

```{r}
p4 <- sg_gdp_long %>%
  # Replace '_' with a space
  mutate(Sector = str_replace_all(Sector, "[[:punct:]]", " ")) %>%
  # Capitalise first letter of each word
  mutate(Sector = str_to_title(Sector)) %>%
  ggplot() +
  geom_line(aes(x = date,
                y = Sector_GDP,
                colour = Sector,
                group = 1,
                # Format legend name for plotly
                text = paste('Date:', date,
                             '<br>Sector:', Sector, 
                             '<br>Sector GDP:', Sector_GDP
                )),
            alpha = 0.6) +
  theme(legend.position = "none") +
  facet_wrap(~ industry,
             nrow = 3,
             scales = "free_y") +
  ggtitle("Quartely GDP values for each Sector")

plotly::ggplotly(p4,
                 tooltip = "text")
```

## Interactive Table: GDP Trends

```{r tb2-wrangle}
tb2 <- sg_gdp_long %>%
  # Remove all '_' 
  mutate(Sector = str_replace_all(Sector, "[[:punct:]]", " ")) %>%
  mutate(Sector = str_to_title(Sector)) %>%
  arrange(Sector)

# Take subset of industrial GDP changes
## Join for quarterly GDP changes for industry


tb2 <- tb2 %>%
  inner_join(ind_summary, by = c("date", "industry")) %>%
  rename(industry_total = total) %>%
  select(-qtr)

# Set year to year-quarter variable - Better for eyes
tb2$date <- zoo::as.yearqtr(tb2$date, format = "%Y-%m-%d")

# Rearrange table
tb2 <- tb2 %>%
  select(date, 
         Sector, 
         Sector_GDP, 
         industry, 
         goods_producing_industries,
         services_producing_industries,
         gdp_at_current_market_prices
         )

# add relative changes in sector and industry 
tb2 <- tb2 %>%
  group_by(Sector) %>%
  mutate(lag_sector_GDP = lag(Sector_GDP, k =1)) %>%
  ungroup() %>%
  group_by(industry) %>%
  mutate(lag_industry_GDP = lag(gdp_at_current_market_prices, k =1))  %>%
  mutate(sector_GDP_change = round((Sector_GDP - lag_sector_GDP)/lag_sector_GDP, digits = 4), 
         industry_GDP_change = round((gdp_at_current_market_prices - lag_industry_GDP)/lag_industry_GDP, digits = 4)) %>%
  select(-c(lag_sector_GDP,
            lag_industry_GDP)) %>%
  relocate(sector_GDP_change, .before = industry) %>%
  relocate(industry, .before = Sector_GDP) %>%
  rename(total_GDP = gdp_at_current_market_prices)

# Replace inf, NaN and NA values with zero
tb2$sector_GDP_change[!is.finite(tb2$sector_GDP_change)] <- 0
tb2$industry_GDP_change[!is.finite(tb2$industry_GDP_change)] <- 0

# Convert date to character to allow DT::datatable to read
tb2 <- tb2 %>%
  mutate(date = as.character(date))
```

```{r}
tb2_interactive <- DT::datatable(tb2,
              colnames = c('Date Recorded', 
                           'Sector', 
                           'Sector\'s Industry',
                           'Sector\'s GDP',
                           'Sector\'s Quarterly GDP Change',
                           'Goods Industry\'s GDP',
                           'Services Industry\'s GDP',
                           'Total GDP',
                           'Industry\'s Quarterly GDP Change'
                           ),
              filter = 'top',
              rownames = FALSE,
              extensions = c("Scroller", 
                             "FixedColumns"), # Make columns visible when scrolling
              options = list(pageLength = 15,
                             scrollY = "200px",
                             scrollX = TRUE)) %>%
  # Format proportion to percentage
  formatPercentage(c('sector_GDP_change', 'industry_GDP_change'), 
                   digits = 2) 

# Frame table
widgetframe::frameWidget(tb2_interactive)
```

\clearpage

# References



