---
title: Shiny App Self-Review
author: Brendi Ang
date: '2020-10-15'
slug: shiny-app-self-review
categories:
  - R
tags:
  - R Markdown
subtitle: ''
summary: ''
authors: []
lastmod: '2020-10-15T12:31:07+11:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
bibliography: [references.bib]
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE)

# Load library
library(tidyverse)
library(htmltools)
library(widgetframe)
library(DT)

# Read-in data
us_state_tests <- read_csv("us_state_tests.csv")
```

## The Shiny Application

The interactive web application was created to provide key data visualisations and allow users to explore Covid-19 trends across the U.S states. The interactive features allow for a more personalised user experience, to explore as little or as much of the analysed data. In all tabs, the user can make use of the slider input and drop-down list to manoeuvre between dates and states across the United States to visualise how the Covid-19 trajectories changes with these inputs.

Due to the size of this application, I do not have enough instant size to publish this app on ShinyApps.io with my free account. For this reason, the link to my Github repository for the Shiny app is provided [here](https://github.com/etc5523-2020/shiny-assessment-BrendiA).


## Functions of the Shiny Application

### Tab 1: United States COVID-19 Cases (per 100,000) by State

<iframe width="720" height="400" src="../vid/tab1_example.mp4" frameborder="0" allowfullscreen>

</iframe>

The dataset used for this section of the application originates from the NY Times (@albert_sun_2020_3981451). This component provides the confirmed Covid-19 cases, per 100,000 population reported by the state on a given date. The total confirmed cases are plotted on the U.S state map, and is made with a sequential colour scale to establish the geographical location of the states with the highest total confirmed cases. This trend is also plotted on a histogram on the sidebar panel, as the comparison between states can be better articulated since people are generally better at comparing lengths on a common scale as compared to shading or colour saturations (@cleveland1984graphical).

### Tab 2: Positive Results in Covid-19 testing

<iframe width="720" height="400" src="../vid/tab2_example.mp4" frameborder="0" allowfullscreen>

</iframe>

The dataset for this component originates from the *covidtracking.com* (@covidtracking). Covid-19 testing promotes early detection of community-acquired Covid-19 cases, which can support public health response to the pandemic such as the allocation of resources for the more vulnerable regions. This section consists of two components. The first tab measures the percentage of Covid-19 tests returning positive on a given date, and is charted on the US state map. Correspondingly, the second segment depicts the aggregated test results as of the chosen date selected by the user. Not to mention, occasional anomalies (e.g. significant drop in tests or 100% positive test results) can be observed and is generally due to case definitions and non-reporting.

# Self Review

## What Was Done Well?

The layout of the app was structured well. For instance, I applied `fluidRow()` to manually divide user interface into grids to my liking and `tabsetPanel()` to combine panels into a stack. Further, I customised the texts in the UI with `shing::tags$` functions to build specific HTML tags, it took me a while to figure out the different HTML tags!

Next, reactive programming was done fairly well. To emphasise, due to the long running-tasks embedded in this app, it was not ideal to wait for the app to compute every reactive expression each time the user changes the input. For this reason, I applied `observeEvent()` function to delay the computation of this reactive expression only until the action button is clicked.

## Areas of Improvement

Debugging shiny applications is something that I need to work on. Firstly, running reactive codes was problematic for me as I was accustomed to the usual R markdown file system. For this assignment, I created a new Rmd file to make reference to when I ran into errors, which was not the best practice. I should utilise tools such as reactive graphs, log statements and breakpoints.

Organising large data files to optimise the speed of the Shiny application was another aspect that I needed to work on. In this case, the joined data set (Covid-19 cases and Map data) consisted of more than 2 million observations. For example, I could have used `geom_sf` or `leaflet()` map to plot the map instead of `geom_polygon` to reduce the object size. By doing it this way, I may not have to join all the map coordinates with the covid-19 data since I could overlay the map with the variables in the Covid-19 data.

## Plausible Things I'd like to add

In retrospect, I would add a plot or a new tab that showcase Covid-19 trends over time. In my current Shiny app, Covid-19 patterns were only shown at a point in time, and it may be difficult to compare the trends across countries over time to better understand the data.

I would also like to add some technical details to the app:

-   **Plotly Input Events and multiple reactive expressions**: Access plotly input events after the map output is generated. For instance, clicking on the state(s) on the U.S. map automatically updates the bar graph to compare the trends of selected states.

-   **Cancel long-running tasks while loading:** Due to the large amount of data, it takes about 7 seconds or so to load when running locally, and users cannot do anything until the function finish running. I want to add a 'cancel button' so users can cancel the computation of reactive expression at any time.

-   **Pre-populate or download shiny input based on URL parameters** As the Covid-19 datasets are updated regularly, I would like to add a download link for users to download the data directly from the source (e.g. GitHub) to provide a complete and up-to-date data of the ongoing outbreak in the Shiny application.

## Limitations of Data/Technology

```{r dt-table, out.height = "300%"}
tests_grades <- us_state_tests %>%
  filter(date == "2020-10-07",
         !is.na(data_quality)) %>%
  select(state, data_quality) %>%
  distinct(state, data_quality) %>%
  arrange(state) %>%
  rename(State = state,
         `Data Quality` = data_quality)

dt <- tests_grades %>% 
  DT::datatable(
    class = 'cell-border stripe',
    rownames = FALSE,
    escape = FALSE,
    options = list(pageLength = 5,
                   autoWidth = TRUE,
                   scrollY = "400px"),
    caption = htmltools::tags$caption(style = 'caption-side: top; text-align: center; color:black; font-size:150% ;','State Grading System for Traceability of Covid-19 Data') )

# Frame table
widgetframe::frameWidget(dt)
```
       
      
     
A limitation to this Shiny App is the credibility of the data. Due to the various state jurisdictions, case definitions and reporting requirements of each U.S state, the reported values may be subjected to significant variations. To illustrate, in the data retrieved from *covidtracking.com*, numerous US states did not report the negative results of the covid-19 testing. The 'state grading system' (shown in the table below) reveals the data completeness of the total Covid-19 testing in each state (breakdown and criteria for the grades can be found  [here](https://docs.google.com/spreadsheets/u/1/d/e/2PACX-1vRL2zG1o-qj9l2sl19d1lj1oHd6WbkJ0ukFwN04a_ms_ANUdgxTMpI7AF-gbQzwOSreJUDx6PEK7Vnq/pubhtml)). 


# R Packages Used

The following packages used in the coding script:

- tidyverse package (@tidyverse)

- htmltools package (@htmltools)

- widgetframe package (@widgetframe)

- DT package (@DT)

# References
