---
title: A short Story of COVID-19 in South Africa
author: Brendi Ang
date: '2020-09-17'
slug: a-short-story-of-covid-19-in-south-africa
categories:
  - R
tags:
  - R Markdown
  - Academic
subtitle: ''
summary: ''
authors: []
lastmod: '2020-09-17T01:32:24+10:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE, fig.height = 7.5, fig.width = 6)

# Load library
library(tidyverse)
library(coronavirus) # devtools::install_github("RamiKrispin/coronavirus")
library(rnaturalearth) # devtools::install_github("ropenscilabs/rnaturalearth")
library(stringr)
library(janitor)
library(crosstalk)
library(plotly)
library(kableExtra)
library(DT)
library(htmlwidgets)
```

# Introduction

This blog will analyse the COVID-19 cases of South Africa. Relative to the previous blog of COVID-19 cases in Singapore South Africa's economic status, infrastructure, education levels, are generally inferior to that of Singapore as they are a third world country. Furthermore, South Africa's population is approximately ten folds of that of Singapore.

In examining the COVID-19 cases in South Africa, I will break down the cases into distinct provinces of South Africa to verify if a bigger population is associated with a larger number of confirmed cases, deaths and recovered cases.


```{r tb1-interactive}
library(coronavirus)


tb1 <- coronavirus %>%
  pivot_wider(names_from = type,
              values_from = cases) %>%
  group_by(country) %>%
  summarise(Confirmed = sum(confirmed),
            Deaths = sum(death),
            Recovered = sum(recovered)) %>%
  arrange(desc(Confirmed)) %>%
  top_n(100) 

# Set thousand separator
tb1$Confirmed <- scales::comma(tb1$Confirmed , accuracy = 1000)
tb1$Deaths <- scales::comma(tb1$Deaths , accuracy = 1000)
tb1$Recovered <- scales::comma(tb1$Recovered , accuracy = 1000)


# Interactive table
tb1_interactive <- DT::datatable(tb1,
              colnames = c("Country",
                           "Total Confirmed Cases",
                           "Total Deaths",
                           "Total Recovered"
                           ),
              filter = 'top',
              rownames = TRUE,
              extensions = c("Scroller",
                             "FixedColumns"), # Make columns visible when scrolling
              options = list(pageLength = 10,
                             scrollY = "200px",
                             scrollX = TRUE)) %>%
  formatStyle('country',
              target = 'row',
              backgroundColor = styleEqual("South Africa", '#cd5c5c'))

# Frame table
widgetframe::frameWidget(tb1_interactive)
```

The table above shows the top 100 countries with the highest confirmed cases. As of today, South Africa ranked 6th in confirmed cases in the world. 

# Getting the Data

To analyse the COVID-19 cases, I extracted the daily summary of COVID-19 cases in the `coronavirus` package. This package contains the raw data in a tidy format, summarising the daily cases by state/province and is readily available in this [GitHub Link](https://github.com/RamiKrispin/coronavirus). Each observation represents the total number of cases recorded in the state/province, where each cases are identified by their case type, spanning from 2020-01-23 and it updated regularly. It was really convenient to use this data set as no data wrangling had to be done.

Next, I downloaded the population data and cumulative confirmed, deaths and recovered cases from this [GitHub Link](https://github.com/RamiKrispin/coronavirus) (@marivate2020use). The cleaning process included using regular expressions to manipulate character strings, and relational joins to join data sets with matching observations. The cleaned data set was in the same format with that of the `coronavirus` dataset.

```{r province-data}
# Cumulative confirmed cases by province
prov_confirmed <- readr::read_csv("https://raw.githubusercontent.com/dsfsi/covid19za/master/data/covid19za_provincial_cumulative_timeline_confirmed.csv")

# Cumulative death cases by province
prov_deaths <- readr::read_csv("https://raw.githubusercontent.com/dsfsi/covid19za/master/data/covid19za_provincial_cumulative_timeline_deaths.csv")

# Cumulative recovered cases by province
prov_recovered <- readr::read_csv("https://raw.githubusercontent.com/dsfsi/covid19za/master/data/covid19za_provincial_cumulative_timeline_recoveries.csv")

# Provincial Population Estimates
prov_popests <- readr::read_csv("https://raw.githubusercontent.com/dsfsi/covid19za/master/data/official_stats/statssa_population_midyear_provincially_2019.csv_") 
```

```{r clean-province-population}
# Clean provincial population estimates data
prov_popests <- prov_popests %>%
  # Remove spaces in variables
  janitor::clean_names() %>%
  # convert character strings to lower case
  mutate(province = tolower(province)) %>%
  # Replace spaces and dashes with "_"
  mutate(province = str_replace(province, " ", "_"),
         province = str_replace(province, "-", "_")) %>%
  rename(pop_ests = population_estimate,
         prop_of_total = percent_of_total_population)
```

```{r clean-province-cases}
# Cleaning function for provincial cases, converts df to long format
clean_prov <- function(df, case_type) {
  df %>%
  rename(eastern_cape = EC,
         free_state = FS,
         gauteng = GP,
         kwazulu_natal = KZN,
         limpopo = LP,
         mpumalanga = MP,
         northern_cape = NC,
         north_west = NW,
         western_cape = WC) %>%
  mutate(date = as.Date(date, format = "%d-%m-%Y")) %>%
  select(-c(UNKNOWN, 
            source, 
            YYYYMMDD, 
            total)) %>% # Will use coronavirus package for total cases
    pivot_longer(cols = eastern_cape:western_cape,
                 names_to = "province",
                 values_to = case_type) 
}

# Clean data frames
prov_confirmed <- clean_prov(prov_confirmed, "confirmed")
prov_deaths <- clean_prov(prov_deaths, "deaths")
prov_recovered <- clean_prov(prov_recovered, "recovered")

# Change total to total based on case_type
names(prov_confirmed)[4] <- paste0("total_", "confirmed")
names(prov_deaths)[4] <- paste0("total_", "deaths")
names(prov_recovered)[4] <- paste0("total_", "recovered")
```

```{r join-province}
# Join all province data
prov_all <- 
  full_join(prov_confirmed, prov_deaths, by = c("province", "date")) %>%
  full_join(prov_recovered, by = c("province", "date")) 


# Convert to long format for plotting
prov_all_long <- prov_all %>%
  pivot_longer(cols = c(confirmed, deaths, recovered),
               names_to = "type",
               values_to = "cases") %>%
  # Replace NAs with 0
  mutate(cases = replace_na(cases, 0))

prov_all_long <- prov_all_long %>%
  left_join(prov_popests, by = "province") %>%
  # Remove punctuations and capitalise first letter of every word
  mutate(province = str_replace_all(province, "[[:punct:]]", " ")) %>%
  mutate(province = str_to_title(province),
         type = str_to_title(type)) %>%
  # Uniquely identify each observation with ID for plotting
  mutate(ID = row_number())
```


# Introduction




# COVID-19 Cases in Singapore

## Getting the Data

```{r read-in-data}
# From coronavirus package: long format of covid cases
covid_SA <- coronavirus %>% 
  filter(country == "South Africa",
         date >= "2020-03-05") # Show data from first case recorded onwards
```

# Plot 1

```{r plot1, eval = FALSE}
p1 <- prov_all_long %>% 
  # Uniquely identify each observation with ID
  mutate(ID = row_number()) %>%
  filter(date == "2020-07-08") %>%
ggplot(aes(x = ID,
           y = cases, 
           size = pop_ests,
           colour = province,
           group = date)) +
  geom_point() +
  scale_size(guide = 'none') + # Remove legend for population estimates as it is shown in plotly
  scale_size_continuous(range = c(5, 10)) + # Increases sizes of dot points
  facet_wrap(~ type,
             scales = "free_y") +
  theme_linedraw() +
  theme(legend.position = "bottom",
        axis.text.x.bottom = element_blank(),
        axis.ticks = element_blank(),
        rect = element_rect(fill = "#df9797"),
        text = element_text(color = "white", size = 10),
        panel.background = element_rect(fill = "#df9797"),
        axis.text = element_text(color = "white"),
        panel.border = element_rect(color = "black", size = 1.2),
        plot.title = element_text(size = 12, face = "bold", colour = "white"),
        panel.grid = element_blank(),
        legend.title = element_text(color = "white", size = 10), # Change size of text in legend title
        legend.text = element_text(color = "white", size = 10), # Change size of text in legend title
        legend.background = element_rect(fill = "black"), # change legend background 
        legend.key = element_rect(fill = "black") # change legend background 
        ) +
  guides(colour = guide_legend(override.aes = list(size = 3)))
```


```{r}
# Filter to only confirmed cases
confirmed_long <- prov_all_long %>%
  filter(type == "Confirmed")

# Create button attributes 
button_list <- lapply(1:length(unique(confirmed_long$date)), function(x){
  list(method = "restyle",
       args = list("transforms[0].value", (unique(confirmed_long$date))[x]),
       label = (unique(prov_all_long$date))[x])
})

type_list <-  list(
  type = 'dropdown',
  active = -1,
  xanchor = 'center',
  yanchor = "top",
  pad = list('r' = 0, 't' = 10, 'b' = 10),
  x = 0.5,
  y = 1,
  buttons = button_list # assign button_list object to buttons
)

# Hide x-axis
ax <- list(
  title = "",
  zeroline = FALSE,
  showline = FALSE,
  showticklabels = FALSE,
  showgrid = FALSE
)

plot_ly(confirmed_long, 
        x = ~ID,
        y = ~ cases,
        size = ~ pop_ests,
        sizes = c(500, 4000), # Adjust size of population estimates
        color = ~ province,
        colors = ~ c("#5ccdcd", "#cd5c5c", "#5c95cd", "#cd5c95", "#cd955c", "#5cc5c0", "#889977", "#546270", "800080"),
        customdata= ~date,
        type = "scatter",
        mode = "markers",
        # Assign hover info
        text = ~ paste('</br> Province: ', province,
                  '</br> Population: ',pop_ests,
                  '</br> Total Cases: ', cases),
        hoverinfo = "text",
        # Assign filter and target
        transforms = list(
          list(
            type = 'filter',
            target = 'customdata',
            operation = '=',
            value = confirmed_long$date
        )
        )
        ) %>%
  layout(updatemenus = list(type_list),
         xaxis = ax,
         title = "Is a Higher Population Associated with More Confirmed Cases?"
         ) %>%
  # Hide ModeBar 
  config(displayModeBar = FALSE)
```


# Plot 2
```{r plot-2, eval = FALSE}
g2 <- ggplot(prov_all_long) +
  geom_line(aes(x = date,
            y = cases,
            colour = province)) +
  facet_wrap(~ type) +
  theme_linedraw() +
  scale_colour_brewer(palette = "Dark2") +
    scale_x_date(date_breaks = "1 week", 
                 date_labels = "%b %d") +
  theme(legend.position = "bottom",
        rect = element_rect(fill = "#df9797"),
        text = element_text(color = "white", size = 10),
        panel.background = element_rect(fill = "#df9797"),
        axis.text = element_text(color = "white"),
        axis.title.x = element_blank(), # Remove x-axis title
        panel.border = element_rect(color = "black", size = 1.2),
        plot.title = element_text(size = 12, face = "bold", colour = "white"),
        panel.grid = element_blank(),
        legend.title = element_text(color = "white", size = 10), 
        legend.text = element_text(color = "white", size = 10), 
        legend.background = element_rect(fill = "black"), 
        legend.key = element_rect(fill = "black"), 
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5) # Rotate x-axis labels
        ) 
```


# Interactive Plot 2
```{r interactive-plot-2}
require(crosstalk)

prov <- highlight_key(prov_all_long)

gg <- ggplot(prov) +
  geom_line(aes(x = date,
            y = cases,
            colour = province,
            group = province,
            text = paste('</br> Date:', date,
                  '</br> Total Cases:', cases,
                  '</br> Province:', province))) +
  facet_wrap(~ type) +
  theme_linedraw() +
  scale_colour_brewer(palette = "Dark2") +
    scale_x_date(date_breaks = "1 week", 
                 date_labels = "%b %d") +
  theme(legend.position = "none",
        rect = element_rect(fill = "#d37070"),
        text = element_text(color = "white", size = 10),
        panel.background = element_rect(fill = "#d37070"),
        axis.text = element_text(color = "white"),
        axis.title = element_blank(), # Remove x-axis title
        panel.border = element_rect(color = "black", size = 1.2),
        plot.title = element_text(size = 12, face = "bold", colour = "white"),
        panel.grid = element_blank(),
        legend.title = element_text(color = "white", size = 10), 
        legend.text = element_text(color = "white", size = 10), 
        legend.background = element_rect(fill = "black"), 
        legend.key = element_rect(fill = "black"), 
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5) # Rotate x-axis labels
        ) 

filter <- bscols(
  filter_select(id = "ID", 
                label = "Select a Province",
                sharedData = prov,
                group = ~ province),
  ggplotly(gg, tooltip = c("text")) %>%
    config(displayModeBar = F), 
  widths = c(12,12) 
)

bscols(filter)
```




# Acknowledgements

The programming language used to analyse the trends of global astronautical exploration is R (4.0.2) [@R]. And the platform which used for running R language is Rstudio (1.3.1056) [@Rstudio].

Following packages has been included in our Rmd file.

- package tidyverse (1.3.0) [@tidyverse],
- package dplyr (1.0.1) [@dplyr],
- package plotly (4.9.2.1) [@plotly],
- package ggplot2 (3.3.2) [@ggplot2]
- package geomnet (0.3.0) [@geomnet]
- package DT (0.15) [@DT]
- package htmltools (0.5.0) [@htmltools]
- package patchwork (1.0.1) [@patchwork]

# References