---
title: R Package 'covda' Review
author: Brendi
date: '2020-10-28'
slug: r-pkg-review
categories:
  - R
tags:
  - Academic
subtitle: ''
summary: ''
authors: []
lastmod: '2020-10-28T21:40:24+11:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE, fig.show = "hold")

library(tidyverse)
library(knitr)
```

# Overview

In this blog, I will be reviewing Lingling Bai's `covda` R package, which is developed for the ETC5523 Package assignment. The package repository is publicly available on [GitHub](https://github.com/etc5523-2020/r-package-assessment-AntiRosa). In reviewing this R package, I have cloned her package repository and will be utilising a check list adapted from the [ropensci review template](https://devguide.ropensci.org/reviewtemplate.html) for reference, this check list can be found in the appendix of this blog.

## What did I think this package would do? 

The intended purpose of this shiny package is to provide an embedded interactive Shiny application within the package that allow users to explore a Covid-19 related data. To be more specific, the R package should have an exported function `launch_app()` which load the aforementioned Shiny application; along with some functions that refactors the user interface or server-side functions within the application.

## Did it do it?

According to the `README` file, the goal of this package is to analyse Covid-19 cases across countries. This package may be installed via the code:

```{r, echo = TRUE, eval = FALSE}
devtools::install_github("etc5523-2020/r-package-assessment-AntiRosa")
```

When I tried running the code above, it could not run. Accordingly, I looked to examine the package's repository. In trying to rectifying this error, I learnt that the package could not be built/installed (*i.e.* I ran into some errors when I ran the code `devtools::install()` in the repository). Furthermore, there were a few examples of the codes found in the README file were not related to a Covid-19 dataset. 

<head>
<style>

.img-container {
  float: left;
  width: 50%;
  padding: 5px;
}

</style>
</head>
<body>

<div class="clearfix">
  <div class="img-container">
  <img src = "/img/dir_tree1.png" width="120%" height="150%"/>
  </div>
  <div class="img-container">
  <img src = "/img/dir_tree2.png" width="120%" height="150%"/>
  </div>
</div>

Two `.Rproj` were enclosed in the working directory, one in the main working directory (`r-package-assessment-AntiRosa.Rproj`) and the other in the `covdata` directory. I attempted to resolve the issues related to building the R package in both .Rproj, but to no avail. However, The `covdata` .Rproj seemed to be more comprehensive than the other R project, and for this reason, I will examine the R package related to `covdata.Rproj`.

## Are there improvements that could be made to the code?

Although the R package was not available for use, the package structure followed the general procedures, such as having a suitable license, storing documentations and functions in the R directory, and holding scripts that generate datasets in the data-raw directory. The package skeleton is shown above in the working directory tree.

There were no exported functions in this R package (*i.e.* No code found in NAMESPACE folder). Some R scripts could be found in the R folder of the package repository, however, none of the scripts were related to a Covid-19 data. Also, some functions such as `key_crop_yields` and `stack-normalise` were not used in the package, but was documented as part of the package.

## Are there improvements that could be made to the shiny app inside the package?

There was no Shiny application embedded in the R package or the package repository, and the `launch_app()` function was not available.

## Is the documentation sufficient to get started?

There were minimal documentation made in this package and no exported functions.

## What I have Learned from Reviewing

-   It is is good practice to run `devtools::check()` from time to time to ensure that functions are working and the package may be loaded properly. Debugging can be a chore if we do not do this!

-   Proper documentation is an important aspect of an R package. Without it, users will not be able to understand what the package does or the functionalities of it.

-   README file should highlight how users can access the package, the requirements needed to load and install the package, and how users can use it.

# Appendix

## Package Review

### Documentation

-   [x] To note, indicates that criteria was met.

The package includes all the following forms of documentation:

-   [x] **Installation instructions:** for the package is found in
    README

-   [ ] **Vignette(s)** demonstrating major functionality that runs
    successfully locally

-   [ ] **Function Documentation:** for all exported functions in R help

-   [ ] **Examples:** for all exported functions in R Help that run
    successfully locally

### Comments

-   README provided instructions on how to install the package.

-   Vignettes could not run locally and no exported functions are
    provided in this package.

### Functionality

-   [ ] **Installation:** Installation succeeds as documented.

-   [ ] **Functionality:** Any functional claims of the software been
    confirmed.

-   [ ] **Automated tests:** Unit tests cover essential functions of the
    package and a reasonable range of inputs and conditions. All tests
    pass on the local machine.

### Comments

-   Code to install package directly from github
    `install_github("etc5523-2020/r-package-assessment-AntiRosa")` could
    not be run.

-   No proper unit tests were conducted.
