---
title: r-pkg-review
author: Brendi
date: '2020-10-28'
slug: r-pkg-review
categories:
  - R
tags:
  - Academic
subtitle: ''
summary: ''
authors: []
lastmod: '2020-10-28T21:40:24+11:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE)

# Load library
library(tidyverse)
```

# Overview

In this blog, I will be reviewing Lingling Bai's `covda` R package, which is developed for the ETC5523 Package assignment. The package repository is publicly available on GitHub [GitHub](https://github.com/etc5523-2020/r-package-assessment-AntiRosa). The intended purpose of this shiny package is to provide an embedded interactive Shiny application within the package that allow users to explore a Covid-19 related data. To be more specific, the R package should have an exported function `launch_app()` which load the aforementioned Shiny application; along with some functions that refactors the user interface or server-side functions within the application. 

In reviewing this R package, I will be utilising a check list adapted from the [ropensci review template](https://devguide.ropensci.org/reviewtemplate.html) for reference, this check list can be found in the appendix of this blog. 

# Functionality of the Package

According to the `README` file, the goal of this package is to analyse Covid-19 cases across countries. 

This package may also be installed via the code:

```{r, eval = FALSE}
devtools::install_github("etc5523-2020/r-package-assessment-AntiRosa") # Could not run
```


## Area of Improvements 









# Appendix

## Package Review

### Documentation

The package includes all the following forms of documentation:

-   [x] **Installation instructions:** for the package is
    found in README

-   [ ] **Vignette(s)** demonstrating major functionality that runs
    successfully locally

-   [ ] **Function Documentation:** for all exported functions in R help

-   [ ] **Examples** for all exported functions in R Help that run
    successfully locally

#### Comments

- README contains clear instructions to install package, however, it should be noted that the package may not be installed on CRAN as stated in the README. 

- No vignettes and exported functions provided in this package.


### Functionality

-   [ ] **Installation:** Installation succeeds as documented.

-   [ ] **Functionality:** Any functional claims of the software been
    confirmed.

-   [ ] **Automated tests:** Unit tests cover essential functions of the
    package and a reasonable range of inputs and conditions. All tests
    pass on the local machine.

#### Comments

- Code to install package directly from github, `install_github("etc5523-2020/r-package-assessment-AntiRosa")` could not be run.

- No unit tests were conducted.


